---
title: "UC2 Sentinel project: evaluation of missing pattern of body mass index (BMI) in flu and covid cohorts of Kaiser Permanente Washington EHR"
author: "Noorie Hyun"
date: "6/19/2024"
output:
 html_document:
        toc: true
        theme: cerulean
        highlight: default
        #code_folding: hide
editor_options: 
        chunk_output_type: console
---


  <style type="text/css">
    h1.title {
      font-size: 26px;
      font-family: "Times New Roman", Times, serif;
      color: DarkBlue;
    }
    h1 { /* Header 1 */
      font-size: 22px;
      font-family: "Times New Roman", Times, serif;
      color: DarkBlue;
    }
    h2 { /* Header 2 */
        font-size: 18px;
        font-family: "Times New Roman", Times, serif;
      color: DarkBlue;
    }
    h3 { /* Header 3 */
      font-size: 14px;
      font-family: "Times New Roman", Times, serif;
      color: DarkBlue;
    }
    pre { /* Code block - determines code spacing between lines */
    font-size: 14px;
    }
  </style>

Output was generated with `r version$version.string`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r  echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(knitr.duplicate.label = 'allow')

library(smdi)
library(naniar) #replace NA
library(haven)
library(dplyr)
library(knitr)
library(tidyverse)
library(arsenal)#comparedf()

```


```{r include=FALSE, message=FALSE, warning=FALSE}

input.dir<-"G:/CTRHS/Sentinel/Innovation_Center/UC2_Shortreed/ANALYSIS/Noorie/Data/"
input.dir2<-"//Groups.ghc.org/data/CTRHS/Sentinel/Innovation_Center/UC2_Shortreed/PROGRAMMING/DATA/PHI_Interim/Pull COD-Covid-19 cohort/"

## read data 

ate<-read_sas(data_file=paste0(input.dir,"r04_ip_ate_dxip.sas7bdat"))%>%data.frame()%>%arrange(PATID)
ate2<-read_sas(data_file=paste0(input.dir,"r04_ip_ate.sas7bdat"))%>%data.frame()%>%arrange(PATID)%>%
  mutate(group=ifelse(exposure==1,"cov_ip_ate_dxip", "flu_ip_ate_dxip"))

covlist<-c("PATID","group",setdiff(names(ate2),names(ate)))

## cod data

flu.cod<-read_sas(data_file=paste0(input.dir,"r04_cod.sas7bdat"))
nrow(flu.cod) #730
n_distinct(flu.cod$PATID) #249

covid.cod<-read_sas(data_file=paste0(input.dir2,"cod_r04.sas7bdat"))
nrow(covid.cod) #574
n_distinct(covid.cod) #574

# 10 subjects with ATE-related deaths 
flu.cod.person<-flu.cod%>%filter(COD_ATE==1)%>%
  group_by(PATID)%>%
  summarise(DaysDiff_index_to_Dth=min(DaysDiff_index_to_Dth))%>%
  mutate(DaysDiff_index_to_Dth = DaysDiff_index_to_Dth +1,
         exposure=0)
  
# 11 subjects with ATE-related deaths 
covid.cod.person<-covid.cod%>%filter(COD_ATE==1)%>%
    group_by(PATID)%>%
    summarise(DaysDiff_index_to_Dth=min(DaysDiff_index_to_Dth),
            deathdt=min(deathdt))%>%
   mutate(DaysDiff_index_to_Dth = DaysDiff_index_to_Dth +1,
         exposure=1)

cod<-bind_rows(flu.cod.person,covid.cod.person)

remove(flu.cod.person,covid.cod.person, flu.cod, covid.cod)

###### Merging datasets #############
# to include variable names without label

race_ethnicity_mapping <- c(
  "1" = "American Indian or Alaska Native",
  "2" = "Asian",
  "3" = "Black or African American",
  "4" = "Native Hawaiian or Other Pacific Islander",
  "5" = "White",
  "0" = "Unknown"
)


ate.ip3<-ate%>%left_join(ate2[,covlist],by=c("PATID","group"))%>%
  left_join(cod,by=c("PATID","exposure"))%>%
  mutate( DaysDiff_index_to_Dth=ifelse(is.na(DaysDiff_index_to_Dth),100000,DaysDiff_index_to_Dth),
    bmi=ifelse(!is.na(Mean_BMI_at_hosp_admit),Mean_BMI_at_hosp_admit,
                    ifelse(NDays_RecentBMI_to_hosp_admit>=1 & NDays_RecentBMI_to_hosp_admit<=90,
                        Median_BMI_prior_hosp_admit, NA)),
         bmi2=ifelse(!is.na(Mean_BMI_at_hosp_admit),Mean_BMI_at_hosp_admit,
                    ifelse(NDays_RecentBMI_to_hosp_admit>=1 & NDays_RecentBMI_to_hosp_admit<=1826,
                        Median_BMI_prior_hosp_admit, NA)),
         group2=ifelse(exposure==1,"COVID","Flu"),
         hosp_yr=year(IndexDt),
        comobodity_cat= cut(COMORBIDSCORE,c(0,1,2,50),right=FALSE),
        raceeth = race_ethnicity_mapping[as.character(RACE)],
        followuptime=ifelse(fup_event==0 & DaysDiff_index_to_Dth<=91, DaysDiff_index_to_Dth, followuptime),
        fup_event=ifelse(fup_event==0 & DaysDiff_index_to_Dth<=91, 1, fup_event),
        fup_time_event=followuptime*fup_event,
        bmi_observed=ifelse(is.na(bmi),0,1),
        bmi2_observed=ifelse(is.na(bmi2),0,1))%>%
  arrange(PATID)


# covariates to remove:
rm_cols <- which(sapply(ate.ip3, function(col)  length(levels(factor(col)))==1)==TRUE)
ate.ip2<-ate.ip3[,-rm_cols]


in.covar<-c("group","Event","Age","sex","HISPANIC","RACE","zip3","COMORBIDSCORE", "NumAV", "NumIP" ,
            setdiff(names(ate.ip2)[(grep("covar",names(ate.ip2)))],c(paste0("covar",seq(63,85)),"covarstrat","covarnum","covar87","covar23") ),"covar70","covar71","COVAR80","COVAR81","COVAR82","COVAR83","fup_dth","bmi","bmi2")

ate.ip<-ate.ip2[,in.covar]

ate.ip<-ate.ip%>%
          replace_with_na(replace = list(zip3 = "Missing"))


## importing labels from SAS data
get_labels <- attr_getter("label")
a<-sapply(ate, get_labels)
data.labels1=data.frame(label=unlist(a))
data.labels1$variable=rownames(data.labels1)


a2<-sapply(ate2, get_labels)
data.labels2=data.frame(label=unlist(a2))
data.labels2$variable=rownames(data.labels2)

data.labels<-data.labels1%>%rename(label1=label)%>%
    full_join(data.labels2%>%rename(label2=label),by="variable")%>%
    mutate(label=ifelse(!is.na(label1) & !is.na(label2),label1,
                        ifelse(is.na(label1) & !is.na(label2),label2,
                               ifelse(!is.na(label1) & is.na(label2),label1,""))))%>%
  select(variable,label)

remove(a,a2,data.labels1,data.labels2)



a<-sapply(ate.ip, get_labels)
#print(data.frame(label=unlist(a)))


```

This  R-markdown result is a part of the Supplementary Material for Pocobelli et al., which is a product of the UC2-sentinel propject. 

# Summarize variables with missing values

* zip3: zip code
* bmi: median of bmi measured within 90 days prior to hospital admission date
* bmi2: median of bmi measured within 5 years prior to hospital admission date

```{r include=TRUE, message=FALSE, warning=FALSE}

smdi_summarize(ate.ip)

```

# visualization of missing patterns of BMI

## By group (Covid vs. Flu)

```{r include=TRUE, message=FALSE, warning=FALSE}

smdi_vis(data=ate.ip,covar=c("bmi","bmi2"),strata="group")

```


## By the primary outcome of arterial thromboembolism (ATE) event (1: event and 0: no event during 90 days after hospitalization) 


```{r include=TRUE, message=FALSE, warning=FALSE}

smdi_vis(data=ate.ip,covar=c("bmi","bmi2"),strata="Event")

```


# Group 1 diagnosis

We computes absolute standardized mean differences (ASMDs) of observed patient characteristics.  If ASMDS is close to 0 (say, <0.1), this implies the missing pattern of BMI does not depend on observed covariates, that is, Missing Completely at Random (MCAR) or Missing Not at Random (MNAR). 


## Absolute Standardized Mean Differences (ASMD) of BMI

### By the Primary ATE outcome

```{r include=TRUE, message=FALSE, warning=FALSE, fig.height=10}

bmi2.ind<-which(names(ate.ip)=="bmi2")
out<-smdi_asmd(data=ate.ip[,-bmi2.ind],covar="bmi" , includeNA=FALSE)

out$bmi$asmd_aggregate
#out$bmi$asmd_table1
out$bmi$asmd_plot

```

## ASMD of BMI2

```{r include=TRUE, message=FALSE, warning=FALSE,fig.height=10}

bmi.ind<-which(names(ate.ip)=="bmi")
out2<-smdi_asmd(data=ate.ip[,-bmi.ind],covar="bmi2",  includeNA=FALSE)

out2$bmi$asmd_aggregate
out2$bmi$asmd_plot

```

## Statisticaltests for the null hypothesis of MCAR of BMI

The null hypothesis for Hotelling’s and LIttle's test that the missing pattern of BMI is MCAR. 

### Two-Sample Hotelling’s T-squared test for missing BMI

```{r include=TRUE, message=FALSE, warning=FALSE}
smdi_hotelling(data=ate.ip[,-bmi2.ind],covar="bmi")
```

### Two-Sample Hotelling’s T-squared test for missing BMI2

```{r include=TRUE, message=FALSE, warning=FALSE}
smdi_hotelling(data=ate.ip[,-bmi.ind],covar="bmi2")

```

## Little Single global chi-square p-value (Little) cross all missing variables WITH the null hypothesis of MCAR 

###  BMI

```{r include=TRUE, message=FALSE, warning=FALSE}
tout<-smdi_little(data=ate.ip[,-bmi2.ind])
tout$p.value
```

###  BMI2

```{r include=TRUE, message=FALSE, warning=FALSE}
tout2<-smdi_little(data=ate.ip[,-bmi.ind])
tout2$p.value
```

# Group 2 diagnosis focusing on MAR

This function trains and fits a random forest classification
model to predict  the missing indicator of BMI given
exposure, outcome, and covariates. If the
resulting area under the receiver operating characteristic
curve (AUC) is meaningfully >0.5, this would give some evidence for MAR against MCAR as being the underlying missingness mechanism.

## BMI
```{r include=TRUE, message=TRUE, warning=FALSE}
output<-smdi_rf(data=ate.ip, covar=c("bmi","bmi2"))
output$bmi$rf_table
```

Covariate importance for prediction of BMI missingness 

```{r include=TRUE, message=TRUE, warning=FALSE,fig.height=10}
output$bmi$rf_plot
```


## BMI2
```{r include=TRUE, message=TRUE, warning=FALSE}
output<-smdi_rf(data=ate.ip, covar=c("bmi","bmi2"))
output$bmi2$rf_table
```


Covariate importance for prediction of BMI2 missingness 

```{r include=TRUE, message=TRUE, warning=FALSE,fig.height=10}
output$bmi2$rf_plot
```

# Group 3 diagnosis

This function  examines the association of the missingness indicator of variable "bmi" and the primary ATE outcome. The function computes both a univariate model and a multi-covariate model adjusted for age, sex, race/ethnicity, dropout status, death status, comorbidity score, number of visits, NUMV. and NUMIP. 

As expected, under a MCAR mechanism, there will be no difference in the outcome between patients with and without a value for BMI.

Under MAR, given that missingness can be sufficiently explained by observed covariates, a spurious association in the univariate model will disappear after adjustment. 

If the missingness followed any MNAR mechanism, an association was observed regardless of adjustment.


```{r include=TRUE, message=TRUE, warning=FALSE}
#Important: don't include variables like ID variables, ZIP codes, dates, etc.

in.covar2<-c("Age","sex","HISPANIC","RACE","group","fup_dth","COMORBIDSCORE", "NumAV", "NumIP" ,paste0("covar",c(seq(1,13),seq(28,36),seq(40,47),49,51,52,58,70,71)))


smdi_outcome(data=ate.ip,
                   covar=c(in.covar2,"bmi"),
                     model=c("logistic"),form_lhs="Event")

smdi_outcome(data=ate.ip,
                   covar=c(in.covar2,"bmi2"),
                     model=c("logistic"),form_lhs="Event")

```


# Reference
Janick Weberpals, Sudha R Raman, Pamela A Shaw, Hana Lee, Bradley G Hammill, Sengwee Toh, John G Connolly, Kimberly J Dandreo, Fang Tian, Wei Liu, Jie Li, José J Hernández-Muñoz, Robert J Glynn, Rishi J Desai, smdi: an R package to perform structural missing data investigations on partially observed confounders in real-world evidence studies, JAMIA Open, Volume 7, Issue 1, April 2024, ooae008,
https://doi.org/10.1093/jamiaopen/ooae008
